---
title: "E. coli abundance comparisons"
author: "Kelsey Jesser"
date: "7/11/2023"
output: pdf_document
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r}
rm(list = ls())
```

#load libraries
```{r}
library(readr)
library(ggplot2)
library(tidyr)
library(dplyr)
library(reshape2)
library(gplots)
library(ggpubr)
library(ggthemr)
library(forcats)
library(ggpmisc)
library(cowplot)
library(ggtext)
```

#set theme
```{r}
ggthemr('fresh')
```

#import data
```{r}
qPCR<-read.csv("qpCR abundances v2.csv")
isolate<-read.csv("Isolate abundances v3.csv")
```

#a few calculations
```{r}
#percent E. coli by qPCR (uida/total bac copies)
qPCR$perc_E.coli<-qPCR$UIDA_copies_ng/qPCR$Total_Bacteria_copies_ng * 100

#convert isolate rel abundances to percents
isolate$MG.abundance.perc<-isolate$MG.abundance *100
```

#filter and melt data for symp/asymp abundance barplots
```{r}
#qPCR
qPCR_melt<-qPCR %>% subset(Include=='y' & pathotype=="pathpos") %>%
  melt(id.vars=c("Sample_ID", "cc", "symp.asymp", "pathotype", "UIDA_stdev"), variable.name=("measurement"), value.name="value")
qPCR_melt$value<-as.numeric(qPCR_melt$value)

#isolate
isolate_melt<-isolate %>% subset(Include=="yes") %>%
  melt(id.vars=c("E..coli.isolate.ID", "Sample_ID", "cc", "symp.asymp", "Pathotype.match", "Rotavirus."), variable.name="measurement", valu.name="value")%>%
  subset(measurement=="MG.abundance.perc")
isolate_melt$value<-as.numeric(isolate_melt$value)
```

#qPCR plots
```{r}
#uida (total E. coli target)
#barplot
df_uida<-qPCR_melt %>% 
  subset(measurement=="UIDA_copies_ng") %>%
  drop_na("value")

df_uida.mean = df_uida %>% 
  group_by(symp.asymp) %>% 
  mutate(mean = mean(value))

uida_bar<-qPCR_melt %>% 
  subset(measurement=="UIDA_copies_ng") %>%
  drop_na("value")%>%
  arrange(value) %>%
  mutate(Sample_ID=fct_reorder(Sample_ID, value)) %>%
  mutate(Sample_ID=fct_reorder(Sample_ID, symp.asymp)) %>%
  ggplot(aes(fill=symp.asymp, y=value, x=Sample_ID))+
    geom_bar(stat="identity")+
    scale_fill_manual(values=c("Symptomatic"="coral2", "Asymptomatic"="steelblue3"))+
    ggtitle("E. coli qPCR abundance")+
    ylab("E. coli abundance (uida copies/ng)")+
    xlab(NULL)+
    geom_errorbar(data=df_uida.mean, aes(Sample_ID, ymax = mean, ymin = mean), size=0.5, linetype = "longdash", inherit.aes = F, width = 1, color="black")+
    geom_errorbar(aes(ymin=value-UIDA_stdev, ymax=value+UIDA_stdev, width=0.3), color="black")+
    theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), axis.text.x=element_text(angle=90, vjust=0.5, hjust=1), legend.title=element_blank(), panel.border=element_rect(color="black", fill=NA, size=0.5))

#boxplot
uida_box<-qPCR_melt %>% 
  subset(measurement=="UIDA_copies_ng") %>%
  drop_na("value")%>%
  ggplot(aes(fill=symp.asymp, y=value, x=symp.asymp))+
    geom_boxplot()+
    scale_fill_manual(values=c("Symptomatic"="coral2", "Asymptomatic"="steelblue3"))+
    ylab("% abundance")+
    xlab(NULL)+
    #stat_compare_means(method="wilcox.test", label.x=0.53)+
    theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(),  legend.position="none", panel.border=element_rect(color="black", fill=NA, size=0.5), axis.text.x=(element_text(size=8)))

#perc E.coli (uida/total back 16S target)
#barplot
df_perc_E.coli<-qPCR_melt %>% 
  subset(measurement=="perc_E.coli") %>%
  drop_na("value")

df_perc_E.coli.mean = df_perc_E.coli %>% 
  group_by(symp.asymp) %>% 
  mutate(mean = mean(value))

perc_E.coli_bar<-qPCR_melt %>% 
  subset(measurement=="perc_E.coli") %>%
  drop_na("value")%>%
  arrange(value) %>%
  mutate(Sample_ID=fct_reorder(Sample_ID, value)) %>%
  mutate(Sample_ID=fct_reorder(Sample_ID, symp.asymp)) %>%
  ggplot(aes(fill=symp.asymp, y=value, x=Sample_ID))+
    geom_bar(stat="identity")+
    scale_fill_manual(values=c("Symptomatic"="coral2", "Asymptomatic"="steelblue3"), name="DEC infection type")+
    ggtitle(expression(paste(italic("E. coli")," qPCR abundance")))+
    ylab("% abundance")+
    xlab(NULL)+
    geom_errorbar(data=df_perc_E.coli.mean, aes(Sample_ID, ymax = mean, ymin = mean), size=0.5, linetype = "longdash", inherit.aes = F, width = 1, color="black")+
   scale_y_continuous(breaks=seq(0,8,4))+
   theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), axis.text.x=element_text(angle=90, vjust=0.5, hjust=1), panel.border=element_rect(color="black", fill=NA, size=0.5), legend.position="none",)

#boxplot
perc_E.coli_box<-qPCR_melt %>% 
  subset(measurement=="perc_E.coli") %>%
  drop_na("value")%>%
  ggplot(aes(fill=symp.asymp, y=value, x=symp.asymp))+
    geom_boxplot()+
    scale_fill_manual(values=c("Symptomatic"="coral2", "Asymptomatic"="steelblue3"), name="DEC infection type")+
    xlab(NULL)+
    scale_y_continuous(breaks=seq(0,8,4))+
    theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), legend.position="none", panel.border=element_rect(color="black", fill=NA, size=0.5), axis.title.y=element_blank(), axis.text.x=(element_text(size=8)))

uida_plot<- ggdraw() +
    draw_plot(uida_bar) +
    draw_plot(uida_box, x = 0.675, y = 0.55, width = 0.25, height = 0.325)

uida_plot

perc_E.coli_plot<- ggdraw() +
    draw_plot(perc_E.coli_bar) +
    draw_plot(perc_E.coli_box, x = 0.675, y = 0.55, width = 0.25, height = 0.325)

perc_E.coli_plot


  
```

#Isolate abundance plots
```{r}
#isolate abundance
#barplot
df_isolate.mean = isolate_melt %>% 
  group_by(symp.asymp) %>% 
  mutate(mean = mean(value))

isolate_bar<-isolate_melt %>% 
  #arrange(value) %>%
  mutate(Sample_ID=fct_reorder(Sample_ID, value)) %>%
  mutate(Sample_ID=fct_reorder(Sample_ID, symp.asymp)) %>%
  ggplot(aes(fill=symp.asymp, y=value, x=Sample_ID))+
    geom_bar(stat="identity")+
    scale_fill_manual(values=c("Symptomatic"="coral2", "Asymptomatic"="steelblue3"), name="DEC infection type")+
    ggtitle("DEC isolate metagenome abundance")+
    ylab("% abundance")+
    xlab(NULL)+
    geom_errorbar(data=df_isolate.mean, aes(Sample_ID, ymax = mean, ymin = mean), size=0.5, linetype = "longdash", inherit.aes = F, width = 1, color="black")+
    theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),  panel.border=element_rect(color="black", fill=NA, size=0.5), legend.position="none", plot.title=element_text(face="plain"))+
    theme(axis.line.x.bottom=element_line(color="black"))+
    theme(axis.line.y.left=element_line(color="black"))


#boxplot
isolate_box<-isolate_melt %>% 
  ggplot(aes(fill=symp.asymp, y=value, x=symp.asymp))+
    geom_boxplot()+
    scale_fill_manual(values=c("Symptomatic"="coral2", "Asymptomatic"="steelblue3"), name="DEC infection type")+
    xlab(NULL)+
    theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), legend.position="none", panel.border=element_rect(color="black", fill=NA, size=0.5), axis.title.y=element_blank(), axis.text.x=(element_text(size=8)))+
    theme(axis.line = element_line(color="black", size =.1), axis.ticks=element_line(color="black"))

isolate_plot<-ggdraw() +
    draw_plot(isolate_bar) +
    draw_plot(isolate_box, x = 0.675, y = 0.55, width = 0.25, height = 0.325)

isolate_plot

```


#qPCR stats
```{r}
#Total E. coli (uida gene copies) symp.asymp comparison
qPCR_melt_symp_uida<-qPCR_melt %>% 
  subset(symp.asymp=="Symptomatic" & measurement=="UIDA_copies_ng") %>%
  drop_na("value")
qPCR_melt_asymp_uida<-qPCR_melt %>%
  subset(symp.asymp=="Asymptomatic" & measurement=="UIDA_copies_ng") %>%
  drop_na("value")

wilcox.test(qPCR_melt_symp_uida$value, qPCR_melt_asymp_uida$value)

#Percent E. coli symp.asymp comparison
qPCR_melt_symp_perc<-qPCR_melt %>%
  subset(symp.asymp=="Symptomatic" & measurement=="perc_E.coli") %>%
  drop_na("value")

#Percent E. coli symp.asymp comparison
qPCR_melt_asymp_perc<-qPCR_melt %>%
  subset(symp.asymp=="Asymptomatic" & measurement=="perc_E.coli") %>%
  drop_na("value")

wilcox.test(qPCR_melt_symp_perc$value, qPCR_melt_asymp_perc$value)

```

#isolate stats
```{r}
#isolate
isolate_melt_symp<-isolate_melt %>% 
  subset(symp.asymp=="Symptomatic") %>%
  drop_na("value")
isolate_melt_asymp<-isolate_melt %>%
  subset(symp.asymp=="Asymptomatic") %>%
  drop_na("value")

wilcox.test(isolate_melt_symp$value, isolate_melt_asymp$value)
```

#qPCR and isolate metagenome abundance comparisons
```{r}
#join abundance data
isolate_qPCR<- left_join(isolate, qPCR, by=c("Sample_ID"))

#formula
formula <- y ~ x 

#percent E. coli
isolate_qPCR_scat_perc<-ggplot(isolate_qPCR, aes(x=perc_E.coli, y=MG.abundance.perc))+
  geom_point(size=3)+
  xlab(expression(paste(italic("E. coli")," qPCR abundance (%)")))+
  ylab("DEC isolate metagenome abundance (%)")+
  geom_smooth(method=lm, linetype="dashed", color="black", se=FALSE)+
  stat_poly_eq(aes(label = paste(..rr.label..)), formula = formula, parse = TRUE, label.y.npc=.99, label.x.npc=0.03)+
  stat_fit_glance(method = 'lm', method.args = list(formula = formula), geom = 'text', aes(label = paste("p = ", signif(..p.value.., digits = 2), sep = "")))+
  theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), legend.position="none", panel.border=element_rect(color="black", fill=NA, size=0.5))
isolate_qPCR_scat_perc
```
#extract legend
```{r}
isolate_legend<-isolate_melt %>% 
  #arrange(value) %>%
  mutate(Sample_ID=fct_reorder(Sample_ID, value)) %>%
  mutate(Sample_ID=fct_reorder(Sample_ID, symp.asymp)) %>%
  ggplot(aes(fill=symp.asymp, y=value, x=Sample_ID))+
    geom_bar(stat="identity")+
    scale_fill_manual(values=c("Symptomatic"="coral2", "Asymptomatic"="steelblue3"), name="DEC infection type", labels=c("Symptomatic DEC infections", "Asympotmatic DEC infections"))+
    #scale_fill_discrete(labels=c("Symptomatic DEC infections"="coral2", "Asymptomatic DEC infections"="steelblue3"))+
    ggtitle("DEC isolate metagenome abundance")+
    ylab("% abundance")+
    xlab(NULL)+
    geom_errorbar(data=df_isolate.mean, aes(Sample_ID, ymax = mean, ymin = mean), size=0.5, linetype = "longdash", inherit.aes = F, width = 1, color="black")+
    theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),  panel.border=element_rect(color="black", fill=NA, size=0.5), legend.position="bottom",legend.title=element_blank())

legend<-get_legend(isolate_legend)
l<-as_ggplot(legend)
l
```


#multiplots
```{r}
#barplots
cowplot::plot_grid(isolate_bar, uida_bar, perc_E.coli_bar, ncol=2)

#boxplots
cowplot::plot_grid(isolate_box, uida_box, perc_E.coli_box, ncol=2)

#comparisons
isolate_qPCR_scat_perc

#inset multiplot
p<-cowplot::plot_grid(isolate_plot, 
             perc_E.coli_plot,
             l,
             ncol=1, align="hv", 
             labels=c("A", "B", ""), 
             label_size=20, 
             rel_heights = c(1, 1, 0.25), 
             rel_widths=c(1, .9, 1))
q<-cowplot::plot_grid(p)+
    draw_text("Wilcoxon, p=0.0049", x = 0.1725, y = .925, size = 10)+
    draw_text("Wilcoxon, p=0.11", x=0.175, y=0.475, size = 10)
q
```

#nicely formatted isolate barplot
```{r}
isolate_bar<-isolate_melt %>% 
  #arrange(value) %>%
  mutate(Sample_ID=fct_reorder(Sample_ID, value)) %>%
  mutate(Sample_ID=fct_reorder(Sample_ID, symp.asymp)) %>%
  ggplot(aes(fill=symp.asymp, y=value, x=Sample_ID))+
    geom_bar(stat="identity")+
    scale_fill_manual(values=c("Symptomatic"="coral2", "Asymptomatic"="steelblue3"), labels=c("Asymptomatic DEC infections", "Symptomatic DEC infections"))+
    #ggtitle("DEC isolate metagenome abundance")+
    ylab("% abundance")+
    xlab(NULL)+
    geom_errorbar(data=df_isolate.mean, aes(Sample_ID, ymax = mean, ymin = mean), size=0.5, linetype = "longdash", inherit.aes = F, width = 1, color="black")+
    scale_y_continuous(expand=c(0,0), limits=c(0,60), breaks=seq(0, 60, by=25))+
    theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),  panel.border=element_rect(color="black", fill=NA, size=0.5), legend.position="bottom", plot.title=element_text(face="plain"), legend.title= element_blank())+
    theme(axis.line = element_line(color="black", size =.1), axis.ticks=element_line(color="black"))
  
isolate_plot<-ggdraw() +
    draw_plot(isolate_bar) +
    draw_plot(isolate_box, x = 0.675, y = 0.625, width = 0.25, height = 0.325)

isolate_plot<-cowplot::plot_grid(isolate_plot)+draw_text("Wilcoxon, p=0.0049", x = 0.17, y = .95, size = 10)
isolate_plot
```

#Subset metagenome isolate abundances for ages 0-5y
```{r}
ages<-read.csv("ages_metadata_final.csv")%>%
  select(c(Alternate.Sample.ID, Age.years, Age.months))

isolate_melt_age<-isolate_melt%>%
  dplyr::rename(Alternate.Sample.ID=Sample_ID)%>%
  left_join(ages, by="Alternate.Sample.ID")

isolate_melt_0to5<-isolate_melt_age%>%
  mutate(Age.years=as.numeric(Age.years))%>%
  subset(Age.years<=4)%>%
  subset(Age.years>=0)
```

#Compare isolate abundances in metagenomes for participants aged <5y
```{r}
#isolate abundance
#barplot
df_isolate.mean = isolate_melt_0to5 %>% 
  group_by(symp.asymp) %>% 
  mutate(mean = mean(value))

isolate_bar_0to5<-isolate_melt_0to5 %>% 
  arrange(value) %>%
  mutate(Alternate.Sample.ID=fct_reorder(Alternate.Sample.ID, value)) %>%
  mutate(Alternate.Sample.ID=fct_reorder(Alternate.Sample.ID, symp.asymp)) %>%
  ggplot(aes(fill=symp.asymp, y=value, x=Alternate.Sample.ID))+
    geom_bar(stat="identity")+
    scale_fill_manual(values=c("Symptomatic"="coral2", "Asymptomatic"="steelblue3"), name="DEC infection type")+
    #ggtitle("DEC isolate metagenome abundance in children ages 0-5y")+
    ylab("% abundance")+
    xlab(NULL)+
    geom_errorbar(data=df_isolate.mean, aes(Alternate.Sample.ID, ymax = mean, ymin = mean), size=0.5, linetype = "longdash", inherit.aes = F, width = 1, color="black")+
    theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), axis.text.x=element_text(angle=90, vjust=0.5, hjust=1),  panel.border=element_rect(color="black", fill=NA, size=0.5), legend.position="bottom", plot.title=element_text(face="plain"))+
    theme(axis.line.x.bottom=element_line(color="black"))+
    theme(axis.line.y.left=element_line(color="black"))

#boxplot
isolate_box_0to5<-isolate_melt_0to5 %>% 
  ggplot(aes(fill=symp.asymp, y=value, x=symp.asymp))+
    geom_boxplot()+
    scale_fill_manual(values=c("Symptomatic"="coral2", "Asymptomatic"="steelblue3"), name="DEC infection type")+
    xlab(NULL)+
    theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(), legend.position="none", panel.border=element_rect(color="black", fill=NA, size=0.5), axis.title.y=element_blank(), axis.text.x=(element_text(size=8)))+
    theme(axis.line = element_line(color="black", size =.1), axis.ticks=element_line(color="black"))

isolate_plot_0to5<-ggdraw() +
    draw_plot(isolate_bar_0to5) +
    draw_plot(isolate_box_0to5, x = 0.65, y = 0.65, width = 0.25, height = 0.3)
isolate_plot_0to5

#stats
isolate_melt_symp_0to5<-isolate_melt_0to5 %>% 
  subset(symp.asymp=="Symptomatic") %>%
  drop_na("value")
isolate_melt_asymp_0to5<-isolate_melt_0to5 %>%
  subset(symp.asymp=="Asymptomatic") %>%
  drop_na("value")

wilcox.test(isolate_melt_symp_0to5$value, isolate_melt_symp_0to5$value)

isolate_plot_0to5<- isolate_plot_0to5+draw_text("Wilcoxon, p=0.02", x = 0.15, y = .95, size = 10)
isolate_plot_0to5
```
